Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-05-31 thesis/data.tex
thesis updates / reorganization of RV modeling

To provide an empirical benchmark for simulations in the next chapters,
we would like to understand and determine the observed properties of
Sculptor and Ursa Minor. In this section, we first review past
measurements of each galaxy and the observed properties of each dwarf
galaxy. We describe the \citet{jensen+2024} Bayesian methodology to
select high probability member stars from \emph{Gaia}. We test the
derived density profiles for Sculptor, Ursa Minor, and Fornax, finding
that \citet{jensen+2024}'s methodology successfully reduces the
background and produces similar results to a simple
background-subtracted density model. We show that Fornax is
approximately represented by a 2-D exponential profile. Instead,
Sculptor and Ursa Minor show an excess of stars relative to an 2-D
exponential in the outer regions. A Plummer profile provides a better
fit instead. Finally, we combine various radial velocity samples with
\citet{jensen+2024}'s members to derive systemic velocities and
dispersions for Sculptor and Ursa Minor.

@@

The determination of basic properties of astronomical objects is an
ongoing and challenging endeavour.

\subsection{Gaia}\label{gaia}

Some of the most fundamental properties of astronomical objects are
their position and velocity. Unfortunately, determining positions to
stars is nontrivial, . Additionally, while line-of-sight or radial
velocities (RVs) are easily determined from spectroscopy, the tangental
velocities, perpendicular to RV, are only measurable through proper
motions, typically requiring precise astrometry as well. \emph{Gaia}'s
mission is to produce extremely precise astrometry enabling measurements
of unprecedented accuracy and scale for proper motions, parallaxes, and
magnitudes.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.3x0.75m mirrors situated at the Sun-Earth L2 lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
completing its mission in 2025 (but with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5degrees, Gaia is able to measure
absolute proper motions by comparing the apparent shifts of stars in
different regions of the sky. In addition to precise astrometric
information, \emph{Gaia} measures the magnitude of stars in the very
wide \emph{G} band (330-1050nm), blue and red colours using the blue and
red photometers (BP and RP), and takes low resolution BP-RP spectra and
radial velocity measurements of bright stars (magnitudes \textgreater{}
16?).

@@

While \emph{Gaia} has performed excellently, a few limitations and
systematic uncertainties could become problematic. Gaia does show
systematic variations in its derived proper motions and parallaxes, as
measured by background quasars \citep{lindegren+2021}. These errors
limit the ability to determine absolute proper motions to some degree,
but since we would like to measure if a star is \emph{consistent} with a
dwarf galaxy, these only factor into the uncertainty budget for orbital
motion and are included in J+24. Additionally, while \emph{Gaia} was
designed to be complete down to about a magnitude of \(G=20\),
overcrowding or dense fields can limit this. \emph{Gaia} can only
successfully separate stars at a distance of about \textbf{X} (a few
times the pixel length of \ldots; pixel width is in scanning direction).
For the galaxies Ursa Minor, Fornax, and Sculptor, the highest density
(in FOrnax), is only \textbf{xxx} which appears to not significantly
affect \emph{Gaia}'s survey completeness. We test for brightness affects
in \textbf{REF}.

For our purposes, \emph{Gaia}

\citet[hereafter J+24]{jensen+2024} utilize \emph{Gaia} data with a
Bayesian framework to determine samples of high-probability members for
many Milky Way group dwarf galaxies.

Here, we breifly describe J+24's membership estimation and discuss how
this informs our observational knoledge of each galaxies density
profile. In general, J+24 use a Bayesian framework incorporating proper
motion (PM), colour-magnitude diagram (CMD), and spatial information to
determine the probability that a given star belongs to the satellite.
J+24 extends \citet{MV2020a}; \citet{MV2020b}. \citep[See also similar
work by][]{pace+li2019, battaglia+2022, pace+erkal+li2022, qi+2022}

\subsection{Summary of properties}\label{summary-of-properties}

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{\label{tbl:scl_obs_props}Observed properties of Sculptor. References are: 1. Ricardo R. Muñoz et al. (2018), 2. Tran et al. (2022), 3. McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.18 \pm 0.23\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $9.71\pm0.17\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & XREF\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{\label{tbl:umi_obs_props}Observed properties of Ursa Minor. References are: (1) Ricardo R. Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al. (2020) and Spencer et al. (2018). }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
distance & $70.1 \pm 3.6$ kpc & 2\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\subsection{Data}\label{data}

J+24 select stars initially from Gaia within a 2--4 degree circular
region centred on the dwarf. J+24 only consider stars with:

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), so astrometry
  is high quality.
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance
  (typically near zero; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute RA and Dec proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\),
  less than 10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental
  velocities of \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  Corrected colour excess is within expectations:
  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
  from \citet{riello+2021}. Removes stars with unreliable photometry.
\item
  \(G\) magnitude is between \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\)
  (dereddened magnitudes). Removes very faint stars and stars
  significantly brighter than the tip of the red giant branch (TRGB)
  magnitude plus the distance modulus uncertainty \(\delta {\rm DM}\).
\item
  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
\end{itemize}

Photometry is dereddened using \citet{schlegel+finkbeiner+davis1998}
extinction maps.

J+24 calculate the probability that any star belongs to either the
satellite or the MW background with \begin{equation}
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\end{equation}

where \(f_{\rm sat}\) is the fraction of candidate members in the field,
and \({\cal L}_{\rm sat}\) and \({\cal L}_{\rm bg}\) are the satellite
(sat) and background (bg) likelihoods respectively. The likelihoods are
the product of a spatial, proper motion, and CMD term: \begin{equation}
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\end{equation}

Each likelihood map is normalized over the respective 2D parameter
space. \(f_{\rm sat}\) scales the likelihoods by the abundance of stars
in the satellite compared to the background.

J+24 consider both a one-component and a two-component spatial
likelihood. The one component model is constructed as a single
exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with
\(R_s\) fixed to the value in the table in \citet{MV2020a}.
Additionally, structural uncertainties (for position angle, ellipticity,
and scale radius) are sampled over to construct the final likelihood
map. The two-component model instead adds a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are solved for. Structural
properties are not accounted for. The PM likelihood is a bivariate
gaussian with variance and covariance equal to each star's proper
motions. (The stellar PM uncertainties is assumed and typically is the
dominant PM uncertainty.)

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
Gaia colour uncertainty at each magnitude. The HB is modelled as a
constant magnitude extending blue of the CMD (mean magnitude of
isochone's HB stars and a 0.1 mag width plus the mean colour error). A
likelihood map is constructed by sampling the distance modulus in
addition to the CMD width, taking the maximum of RGB and HB likelihoods.

@@

In J+24, a MCMC simulation is ran using the above total likelihood to
solve \(\mu_{\alpha*}\), \(\mu_\delta\), and \(f_{\rm sat}\). For the
two component models, the amplitude and scale radius of the excented
compopnet is also solved for. The proper motion single component prior
is same as \citet{MV2020a}: a normal distribution with mean 0 and
standard deviation 100 km/s. If 2-component spatial, instead is a
uniform distribution spanning 5\(\sigma\) of single component case w/
systematic uncertainties. Priors on \(f_{\rm sat}\) and \(B\) (if 2
component) are uniform from 0--1. \(R_{\rm outer}\) has a uniform prior
only restricting \(R_{\rm outer} > R_s\). The mode of each parameter
from the MCMC are then used to calculate the final \(P_{\rm sat}\)
values we use here.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to either 0 or 1, so
the choice of probability threshold is only marginally significant.
Additionally, even for a probability cut of 0.2, the purity of the
resulting sample with RV measurements is very high (\textasciitilde90\%,
J+24). Note there is likely a systematic bias in using stars with RV
measurements to measure purity. Fainter stars are less likely to have
been targeted and have poorer astrometry and colour information. We find
no difference in the resulting density distributions when restricting
stars to be brighter than a specific magnitude.

@@

In Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, we
illustrate the resulting samples from the algorithm in the tangent plane
(\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and proper
motion space. For both galaxies, each criteria plays a commensurate role
in sifting out nonmembers. Proper motions are clustered around the dwarf
systemic motion, the CMD is well defined including the horizontal
branch, and stars only within a few \(R_h\) are included. We also mark
stars with consistent radial velocities (RVs) (see below), which trace
each feature albeit more cautiously. We additionally mark the furthest
confirmed RV members of each galaxy, illustrating the large extent of
each galaxy past \(R_h\).

Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the CMD+PM selection as stars satisfying
\begin{equation}
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
\end{equation} These stars are distributed similar to the fiducial
(probable members) sample but with a fainter uniform distribution across
the entire field. This illustrates the approximate background of stars
which may be confused as members. Additionally, since there is no clear
spatial structure in the CMD+PM sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
Gaia. Not shown here, we also try a variety of simpler, absolute cuts
and thresholds, finding no extended structure beyond what is detected in
J+24. (\emph{Is it worth using this to calculate an OOM upper limit on
the density of tidal tails?}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Field stars (satesfying quality criteria) are light grey,
CMD+PM selected stars are blue, probable members (2-component) are
orange, and RV confirmed members are red. \textbf{Top:} Tangent plane
\(\xi, \eta\). We outline in star symbols the two stars from
\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram in
Gaia G versus BP - RP. \textbf{Bottom right:} Proper motion in
declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
(corrected).}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor Selection]{Similar to
Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
members outside of \(3R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor for
different selection criteria, ploted as log surface dencity versus log
elliptical radius. Residuals are with respect to th interpolated
\emph{probable members} density. \emph{Probable members} selects stars
with PSAT \textgreater{} 0.2 considering PM, CMD, and spatial,
\emph{CMD+PM} select stars more likely to be members according to CMD
and PM only, \emph{all} selects any high quality star, and \emph{BG
subtracted} is the background-subtracted density derived from
high-quality stars. We mark the half-light radius ( vertical dashed
line) and the break radius (black arrow,
REF).}\label{fig:scl_observed_profiles}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods.pdf}}
\caption[Ursa Minor density profiles]{Similar to
fig.~\ref{fig:scl_observed_profiles} except for Ursa
Minor.}\label{fig:umi_observed_profiles}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_density_methods.pdf}}
\caption[Fornax density profiles]{Similar to
fig.~\ref{fig:scl_observed_profiles} except for
Fornax.}\label{fig:fornax_observed_profiles}
\end{figure}

@@

\section{Radial velocity modeling}\label{radial-velocity-modeling}

>>>>

To provide an empirical benchmark for simulations in the next chapters,
we aim to understand the observed properties of Sculptor and Ursa Minor.
In this section, we first compile past measurements of each galaxy. We
describe the \citet{jensen+2024} Bayesian methodology to select high
probability member stars from \emph{Gaia}. The derived density profiles
are robust to changes in assumptions. Fornax is well represented by a 2D
exponential profile. Instead, Sculptor and Ursa Minor show an excess of
stars relative to an 2D exponential in the outer regions, better
described my a Plummer profile. Finally, we combine various radial
velocity samples with \citet{jensen+2024}'s members to derive systemic
velocities and dispersions for Sculptor and Ursa Minor.

@@

Since the discovery of dwarf galaxies around the Milky Way,
observational work has attempted to measure and refine the basic
properties of these objects. While the Milky Way's satellites are close
(by extragalactic standards), their low numbers of stars and large areas
on the sky present challenges for observational work. The classical
systems discussed here each extend about 1-2 degrees across the sky
(figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}).

In particular, \emph{Gaia} has provided a wealth of high quality data,
including the distances to nearby stars and proper motions. This allows
for some of the first measurements on the motions for many local group
dwarf galaxies {[}e.g.; \citet{MV2020}{]}. Accurate distances and
velocities are invaluable for understanding the orbital history of
satellites.

High-quality large spectroscopic samples combined with \emph{Gaia} has
opened new windows into understanding the internal kinematics and
history of dwarf galaxies. \textless Discuss battaglia, pace, etc. and
conclusions about dwarf galaxies.\textgreater{}

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: 1. Ricardo R. Muñoz et al. (2018), 2. Tran et al. (2022), 3. McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.3 \pm 0.2\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $9.64\pm0.16\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$R_h$ & $9.79 \pm 0.04$ arcmin & 4\\
$R_{h,inner}$ & $9.46 \pm 0.26$ arcmin & XREF\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Ricardo R. Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al. (2020) and Spencer et al. (2018). }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
distance & $70.1 \pm 3.6$ kpc & 2\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
$R_{h, \rm inner}$ & $11.7\pm0.5$ & \\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{\texorpdfstring{The \emph{Gaia}
mission}{The Gaia mission}}\label{the-gaia-mission}

Some of the most fundamental properties of astronomical objects are
their position and velocity. Unfortunately, determining distances to
stars is nontrivial. Additionally, while line-of-sight or radial
velocities (RVs) are easily determined from spectroscopy, the tangental
velocities, perpendicular to RV, are only measurable through proper
motions, typically requiring precise astrometry as well. \emph{Gaia}'s
mission is to produce extremely precise astrometry enabling measurements
of unprecedented accuracy and scale for proper motions, parallaxes, and
magnitudes.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
completing its mission in 2025 (but with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5degrees, \emph{Gaia} is able to
measure absolute proper motions by comparing the apparent shifts of
stars in different regions of the sky. In addition to precise
astrometric information, \emph{Gaia} measures the magnitude of stars in
the very wide \emph{G} band (330-1050nm), blue and red colours using the
blue and red photometers (BP and RP, 330-680, 640-1050 respectively),
and takes low resolution BP-RP spectra and radial velocity measurements
of bright stars (magnitudes \textless16?).

@@

@@

Here, we briefly describe J+24's membership estimation method. J+24 use
a Bayesian framework incorporating proper motion (PM), colour-magnitude
diagram (CMD), and spatial information to determine the probability that
a given star belongs to the satellite. By accounting for PM in
particular, J+24 produces low contamination samples of candidate member
stars. J+24 extends the algorithm presented in \citet{MV2020a};
\citet{MV2020b} but additionally includes an optional secondary,
extended spatial component to find possible members as far as
\textasciitilde10 half-light radii \(R_h\) from some dwarf galaxies. See
also similar work by \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree circular region centred on the dwarf satisfying:

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegrenXXX}.
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance
  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  Corrected colour excess is within expectations:
  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
  from \citet{riello+2021}. Removes stars with unreliable photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\delta {\rm DM}\).
\item
  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
\end{itemize}

Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
extinction maps.

J+24 define likelihoods \({\cal L}\) representing the probability
density that a star is consistent with either the MW stellar background
(\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
product of a spatial, PM, and CMD term: \begin{equation}
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\end{equation}

Each likelihoods is normalized over their respective 2D parameter space
for both the satellite. Since the likelihoods are normalized,
\(f_{\rm sat}\), representing the fraction of member stars in the field,
controls the relative scaling between the satellite and background
likelihoods. The total likelihood for any star in this model is then
\begin{equation}\protect\phantomsection\label{eq:Ltot}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
\end{equation}

For the satellite's spatial likelihood, J+24 consider both one-component
and a two-component cases. The one component model is constructed as a
single exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)),
with \(R_s\) fixed to the value in the table in \citet{MV2020a}.
Additionally, structural uncertainties (for position angle, ellipticity,
and scale radius) are sampled over to construct the final likelihood
map. The two-component model instead adds a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are solved for. Structural
properties are not accounted for.

The PM likelihood is a bivariate gaussian with variance and covariance
equal to each star's proper motions. J+24 assume the stellar PM errors
are the main source of uncertainty.

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
Gaia colour uncertainty at each magnitude. The HB is modelled as a
constant magnitude extending blue of the CMD (mean magnitude of -2.2, 12
Gyr HB stars and a 0.1 mag width plus the mean colour error). A
likelihood map is constructed by sampling the distance modulus in
addition to the CMD width, taking the maximum of RGB and HB likelihoods.

@@

J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
with likelihood from Eq.~\ref{eq:Ltot}. Priors are weakly informative or
uniform. The proper motion single component prior is same as
\citet{MV2020a}: a normal distribution with mean 0 and standard
deviation \(100\,\kms\). If 2-component spatial, instead is a uniform
distribution spanning 5\(\sigma\) of single component case w/ systematic
uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
\(R_{\rm outer}\) has a uniform prior only restricting
\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
then reported and used to calculate the final \(P_{\rm sat}\) values.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to either 0 or 1, so
the choice of probability threshold is only marginally significant.
Additionally, even for a probability cut of 0.2, the purity of the
resulting sample with RV measurements is very high (\textasciitilde90\%,
J+24). Note there is likely a systematic bias in using stars with RV
measurements to measure purity. Fainter stars are less likely to have
been targeted and have poorer astrometry and colour information. We find
no difference in the resulting density distributions when restricting
stars to be brighter than a specific magnitude.

@@

In
Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
we illustrate the resulting samples from the algorithm in the tangent
plane (\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and
proper motion space. For all galaxies, each criteria plays a
commensurate role in sifting out nonmembers. The CMD is well defined and
probable members only extend a few times the colour uncertainty from the
CMD. In proper motion space, selected stars are within \(\masyr\) from
the systematic proper motion. The stars furthest away in proper motion
space typically have large uncertainties \(\sim 1\masyr\), so most
members are reasonably consistent. In any case, the algorithm does
enable removing many stars which do not have consistent proper motions
with the dwarf. Finally, the spatial likelihood reduces the
probabilities of stars distant from the dwarfs centres. Member stars are
rare outside of \(\sim5R_h\), resulting from both the likelihood
specification and the lower density of background stars consistent with
the CMD and PM of the satellite.

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}
also illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a fainter uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
cuts and thresholds, finding no evidence of other structure in both
\emph{Gaia} and DELVE/UNIONS data.

Finally, we illustrate the location of RV-confirmed members from
sec.~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
brighter than a typical \emph{Gaia} candidate, RV members typically have
more precise PMs. However, the RV members fill about the same area on
the CMD down to the magnitude limit. Additionally, while the number of
stars with RV measurements decreases at large radii, there are still
confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
stars illustrate the extended profiles of each galaxy. Note that for an
exponential profile, 99.95\% of stars should be within \(6R_h\), so the
discovery of any stars beyond \(6R_h\) hints at a deviation from an
exponential profile. These \emph{extratidal} stars hint at something
tidal in nature, and our goal is to explore possible interpretations.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor sample selection]{The selection criteria for Scl
members. Field stars (satisfying quality criteria) are light grey,
CMD+PM selected stars are turqoise (eq.~\ref{eq:sel_cmd_pm}, probable
members (2-component) are blue diamonds, and RV confirmed members are
indigo crosses. We mark the two stars from \citet{sestito+2023a} with
orange-outlined indigo stars. \textbf{Top:} Tangent plane \(\xi, \eta\).
The ellipse represents 3 half-light radii. \textbf{Bottom left:} Colour
magnitude diagram in Gaia G versus BP - RP. We plot the Padova 12Gyr
MH=-1.68 isochone in orange. The black bar in the top left represents
the median colour error. \textbf{Bottom right:} Proper motion in
declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
orange circle represents the \citet{MV2020b} proper motion. The black
cross represents the median proper motion
error.}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.pdf}}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:sculptor_selection} except for
Fornax.}\label{fig:fornax_selection}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor,
Ursa Minor, and Fornax for different selection criteria, ploted as log
surface dencity versus log elliptical radius. Residuals are with respect
to th interpolated \emph{probable members} density. \emph{Probable
members} selects stars with PSAT \textgreater{} 0.2 considering PM, CMD,
and spatial, \emph{CMD+PM} select stars more likely to be members
according to CMD and PM only, \emph{all} selects any high quality star,
and \emph{BG subtracted} is the background-subtracted density derived
from high-quality stars. We mark the half-light radius ( vertical dashed
line) and the break radius (black arrow,
REF).}\label{fig:scl_observed_profiles}
\end{figure}

@@

While \emph{Gaia} has shown excellent performance, some notable
limitations may introduce problems in our interpretation and reliability
of density profiles.

Gaia systematics in proper motions and parallaxes are typically smaller
than the values for sources of magnitudes \(G\in[18,20]\). Since we use
proper motions and parallaxes as general consistency with the dwarf, and
factor in systematic uncertainties in each case, these effects should
not be too significant. However, the systematic proper motion
uncertainties becomes the dominant source of uncertainty in the derived
systemic proper motions of each galaxy (see
tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
